
\thispagestyle{plain}

{\small
\setlength{\parindent}{0pt}%
\raggedright\label{copyright-details}
\setlength{\parskip}{10pt}
{\centering

{\Huge\ccbyncnd}

This work is licensed under a Creative Commons\\
Attribution-NonCommercial-NoDerivs 3.0 Unported Licence.\\
\href{http://creativecommons.org/licenses/by-nc-nd/3.0/}{http://creativecommons.org/licenses/by-nc-nd/3.0/}

}

You are free:

\begin{itemize}
\item to copy, distribute, display and perform the work
\end{itemize}

Under the following conditions:

\begin{itemize}
\item Attribution: You must give the original author credit.
\item Non-Commercial: You may not use this work for commercial purposes.
\item No Derivative Works: You may not alter, transform, or build upon this work.
\end{itemize}

With the understanding that:

\begin{itemize}
\item Waiver: Any of the above conditions can be waived if you get permission from the copyright holder.
\item Public Domain: Where the work or any of its elements is in the public domain under applicable law, that status is in no way affected by the licence.
\item Other Rights: In no way are any of the following rights affected by the licence:
\begin{itemize}
\item Your fair dealing or fair use rights, or other applicable copyright exceptions and limitations;
\item The author's moral rights;
\item Rights other persons may have either in the work itself or in how the work is used, such as publicity or privacy rights.
\end{itemize}
\item Notice: For any reuse or distribution, you must make clear to others the licence terms of this work.
\end{itemize}

Amaravati Publications asserts its moral right to be identified as the
author of this book.

Amaravati Publications requests that you attribute ownership of the work
to Amaravati Publications on copying, distribution, display or
performance of the work.

}

% End of copyright-details.tex
