% vim: foldmethod=marker foldlevel=0 foldtext=FoldText()

\chapterTocDelegatePageNumber
\chapter{Setting in Motion the Wheel of Dhamma}% {{{1

\setTocDelegatedPageNumber
\englishText
\renewcommand{\englishTitle}{Setting in Motion the Wheel of Dhamma}

\begin{leader}
\soloinstr{Solo introduction}

This is the first teaching of the Tathāgata on attaining to unexcelled,
perfect enlightenment.

Here is the perfect turning of the incomparable wheel of Truth,
inestimable wherever it is expounded in the world.

Disclosed here are two extremes, and the Middle Way, with the Four Noble
Truths and the purified knowledge and vision Pointed out by the Lord of
Dhamma.

Let us chant together this Sutta proclaiming the supreme, independent
enlightenment that is widely renowned as ‘The~Turning of the Wheel of
the Dhamma.’

\end{leader}

[Thus have I heard]

Once when the Blessed One was staying in the deer sanctuary at
Isipatana, near Benares, he spoke to the group of five bhikkhus:

‘These two extremes, bhikkhus, should not be followed by one who has
gone forth: sensual indulgence, which is low, coarse, vulgar, ignoble,
and unprofitable; and self-torture, which is painful, ignoble, and
unprofitable.

‘Bhikkhus, by avoiding these two extremes, the Tathāgata has realized
the Middle Way, which gives vision and understanding, which leads to
calm, penetration, enlightenment, to Nibbāna.

‘And what, bhikkhus, is the Middle Way realized by the Tathāgata, which
gives vision and understanding, which leads to calm, penetration,
enlightenment, to Nibbāna?

\chapterTocSubIndentTrue
\chapter{Dhammacakkappavattana Sutta}% {{{1

\paliText
\renewcommand{\paliTitle}{Dhammacakkappavattana Sutta}

\begin{leader}
\soloinstr{Solo introduction}

\begin{solotwochants}
Anuttaraṃ abhisambodhiṃ & sambujjhitvā tathāgato\\
Pathamaṃ yaṃ adesesi & dhammacakkaṃ anuttaraṃ\\
Sammadeva pavattento & loke appativattiyaṃ\\
Yatthākkhātā ubho antā & patipatti ca majjhimā\\
Catūsvāriyasaccesu & visuddhaṃ ñāṇadassanaṃ\\
Desitaṃ dhammarājena & sammāsambodhikittanaṃ\\
Nāmena vissutaṃ suttaṃ & dhammacakkappavattanaṃ\\
Veyyākaraṇapāthena & saṅgītantam bhaṇāma se.\\
\end{solotwochants}
\end{leader}

[Evaṃ me sutaṃ]

Ekaṃ samayaṃ bhagavā bārāṇasiyaṃ viharati isipatane migadāye. tatra kho
bhagavā pañcavaggiye bhikkhū āmantesi:

Dve me, bhikkhave, antā pabbajitena na sevitabbā: yo cāyaṃ kāmesu
kāmasukhallikānuyogo; hīno, gammo, pothujjaniko, anariyo,
anatthasañhito; yo cāyaṃ attakilamathānuyogo; dukkho, anariyo,
anatthasañhito.

Ete te, bhikkhave, ubho ante anupagamma majjhimā paṭipadā tathāgatena
abhisambuddhā cakkhukaraṇī, ñāṇakaraṇī, upasamāya, abhiññāya,
sambodhāya, nibbānāya saṃvattati.

Katamā ca sā, bhikkhave, majjhimā paṭipadā tathāgatena abhisambuddhā
cakkhukaraṇī ñāṇakaraṇī, upasamāya, abhiññāya, sambodhāya, nibbānāya
saṃvattati?

\clearpage

\englishText
\markboth{\englishTitle}{\rightmark}

‘It is just this Noble Eightfold Path, namely:

‘Right View, Right Intention, Right Speech, Right Action, Right
Livelihood, Right Effort, Right Mindfulness, and Right Concentration.

‘Truly, bhikkhus, this Middle Way understood by the Tathāgata produces
vision, produces knowledge, and leads to calm, penetration,
enlightenment, to Nibbāna.

‘This, bhikkhus, is the Noble Truth of dukkha:

‘Birth is dukkha, ageing is dukkha, death is dukkha, grief,
lamentation, pain, sorrow and despair are dukkha, association with the
disliked is dukkha, separation from the liked is dukkha, not to get what
one wants is dukkha. In brief, clinging to the five khandhas is dukkha.

‘This, bhikkhus, is the Noble Truth of the cause of dukkha:

‘The craving which causes rebirth and is bound up with pleasure and
lust, ever seeking fresh delight, now here, now there; namely, craving
for sense pleasure, craving for existence, and craving for annihilation.

‘This, bhikkhus, is the Noble Truth of the cessation of dukkha:

‘The complete cessation, giving up, abandonment of that craving,
complete release from that craving, and complete detachment from it.

‘This, bhikkhus, is the Noble Truth of the way leading to the cessation
of dukkha:

‘Only this Noble Eightfold Path; namely, Right View, Right Intention,
Right Speech, Right Action, Right Livelihood, Right Effort, Right
Mindfulness, and Right Concentration.

‘With the thought, “This is the Noble Truth of dukkha,” there arose in
me, bhikkhus, vision, knowledge, insight, wisdom, light, concerning
things unknown before.

\clearpage

\paliText
\markboth{\paliTitle}{\rightmark}

Ayameva ariyo aṭṭhaṅgiko maggo seyyathīdaṃ:

Sammā-diṭṭhi, sammā-saṅkappo, sammā-vācā, sammā-kammanto, sammā-ājīvo,
sammā-vāyāmo, sammā-sati, sammā-samādhi.

Ayaṃ kho sā, bhikkhave, majjhimā paṭipadā tathāgatena abhisambuddhā
cakkhukaraṇī ñāṇakaraṇī, upasamāya, abhiññāya, sambodhāya, nibbānāya
saṃvattati.

Idaṃ kho pana, bhikkhave, dukkhaṃ ariyasaccaṃ:

Jātipi dukkhā, jarāpi dukkhā, maranampi dukkhaṃ,
soka-parideva-dukkha-domanassupāyāsāpi dukkhā, appiyehi sampayogo
dukkho, piyehi vippayogo dukkho, yam-picchaṃ na labhati tampi dukkhaṃ,
saṅkhittena pañcupādānakkhandā dukkhā.

Idaṃ kho pana, bhikkhave, dukkhasamudayo ariyasaccaṃ:

Yāyaṃ taṇhā ponobbhavikā nandirāgasahagatā tatra tatrābhinandinī
seyyathīdaṃ: kāmataṇhā, bhavataṇhā, vibhavataṇhā.

Idaṃ kho pana, bhikkhave, dukkhanirodho ariyasaccaṃ:

Yo tassā yeva taṇhāya asesavirāganirodho, cāgo, paṭinissaggo, mutti,
anālayo.

Idaṃ kho pana, bhikkhave, dukkhanirodhagāminī paṭipadā ariyasaccaṃ:

Ayameva ariyo aṭṭhaṅgiko maggo seyyathīdam: sammā-diṭṭhi,
sammā-saṅkappo, sammā-vācā, sammā-kammanto, sammā-ājīvo, sammā-vāyāmo,
sammā-sati, sammā-samādhi.

[Idaṃ dukkhaṃ] ariyasaccanti me bhikkhave, pubbe ananussutesu dhammesu
cakkhuṃ udapādi, ñāṇaṃ udapādi, paññā udapādi, vijjā udapādi, āloko
udapādi.

\clearpage

\englishText
\markboth{\englishTitle}{\rightmark}

‘With the thought, “This is the Noble Truth of dukkha, and this dukkha
has to be understood,” there arose in me, bhikkhus, vision, knowledge,
insight, wisdom, light, concerning things unknown before.

‘With the thought, “This is the Noble Truth of dukkha, and this dukkha
has been understood,” there arose in me, bhikkhus, vision, knowledge,
insight, wisdom, light, concerning things unknown before.

‘With the thought, “This is the Noble Truth of the cause of dukkha,”
there arose in me, bhikkhus, vision, knowledge, insight, wisdom, light,
concerning things unknown before.

‘With the thought, “This is the Noble Truth of the cause of dukkha, and
this cause of dukkha has to be abandoned,” there arose in me, bhikkhus,
vision, knowledge, insight, wisdom, light, concerning things unknown
before.

‘With the thought, “This is the Noble Truth of the cause of dukkha, and
this cause of dukkha has been abandoned,” there arose in me, bhikkhus,
vision, knowledge, insight, wisdom, light, concerning things unknown
before.

‘With the thought, “This is the Noble Truth of the cessation of
dukkha,” there arose in me vision, knowledge, insight, wisdom, light,
concerning things unknown before.

‘With the thought, “This is the Noble Truth of the cessation of dukkha,
and this cessation of dukkha has to be realized,” there arose in me
vision, knowledge, insight, wisdom, light, concerning things unknown
before.

‘With the thought, “This is the Noble Truth of the cessation of dukkha,
and this cessation of dukkha has been realized,” there arose in me
vision, knowledge, insight, wisdom, light, concerning things unknown
before.

\clearpage

\paliText
\markboth{\paliTitle}{\rightmark}

Taṃ kho panidaṃ dukkhaṃ ariyasaccaṃ pariññeyyanti me bhikkhave, pubbe
ananussutesu dhammesu cakkhuṃ udapādi, ñāṇaṃ udapādi, paññā udapādi,
vijjā udapādi, āloko udapādi.

Taṃ kho panidaṃ dukkhaṃ ariyasaccaṃ pariññātanti me bhikkhave, pubbe
ananussutesu dhammesu cakkhuṃ udapādi, ñāṇaṃ udapādi, paññā udapādi,
vijjā udapādi, āloko udapādi.

Idaṃ dukkhasamudayo ariyasaccanti me bhikkhave, pubbe ananussutesu
dhammesu cakkhuṃ udapādi, ñāṇaṃ udapādi, paññā udapādi, vijjā udapādi,
āloko udapādi.

Taṃ kho panidaṃ dukkhasamudayo ariyasaccaṃ pahātabbanti me bhikkhave,
pubbe ananussutesu dhammesu cakkhuṃ udapādi, ñāṇaṃ udapādi, paññā
udapādi, vijjā udapādi, āloko udapādi.

Taṃ kho panidaṃ dukkhasamudayo ariyasaccaṃ pahīnanti me bhikkhave, pubbe
ananussutesu dhammesu cakkhuṃ udapādi, ñāṇaṃ udapādi, paññā udapādi,
vijjā udapādi, āloko udapādi.

Idaṃ dukkhanirodho ariyasaccanti me bhikkhave, pubbe ananussutesu
dhammesu cakkhuṃ udapādi, ñāṇaṃ udapādi, paññā udapādi, vijjā udapādi,
āloko udapādi.

Taṃ kho panidaṃ dukkhanirodho ariyasaccaṃ sacchikātabbanti me bhikkhave,
pubbe ananussutesu dhammesu cakkhuṃ udapādi, ñāṇaṃ udapādi, paññā
udapādi, vijjā, udapādi āloko udapādi.

Taṃ kho panidaṃ dukkhanirodho ariyasaccaṃ sacchikatanti me bhikkhave,
pubbe ananussutesu dhammesu cakkhuṃ udapādi, ñāṇaṃ udapādi, paññā
udapādi, vijjā udapādi, āloko udapādi.

\clearpage

\englishText
\markboth{\englishTitle}{\rightmark}

‘With the thought, “This is the Noble Truth of the way leading to the
cessation of dukkha,” there arose in me vision, knowledge, insight,
wisdom, light, concerning things unknown before.

‘With the thought, “This Noble Truth of the way leading to the
cessation of dukkha has to be developed,” there arose in me vision,
knowledge, insight, wisdom, light, concerning things unknown before.

‘With the thought, “This Noble Truth of the way leading to the
cessation of dukkha has been developed,” there arose in me vision,
knowledge, insight, wisdom, light, concerning things unknown before.

‘So long, bhikkhus, as my knowledge and vision of reality regarding
these Four Noble Truths, in their three phases and twelve aspects, was
not fully clear to me, I did not declare to the world of spirits,
demons, and gods, with its seekers and sages, celestial and human
beings, the realization of incomparable, perfect enlightenment.

‘But when, bhikkhus, my knowledge and vision of reality regarding these
Four Noble Truths, in their three phases and twelve aspects, was fully
clear to me, I declared to the world of spirits, demons, and gods, with
its seekers and sages, celestial and human beings, that I had realized
incomparable, perfect enlightenment.

‘Knowledge and vision arose: “Unshakeable is my deliverance; this is
the last birth, there will be no more renewal of being.”\thinspace ’

Thus spoke the Blessed One. Glad at heart, the group of five bhikkhus
approved of the words of the Blessed One.

\clearpage

\paliText
\markboth{\paliTitle}{\rightmark}

Idaṃ dukkhanirodhagāminī patipadā ariyasaccanti me bhikkhave, pubbe
ananussutesu dhammesu cakkhuṃ udapādi, ñāṇaṃ udapādi, paññā udapādi,
vijjā udapādi, āloko udapādi.

Taṃ kho panidaṃ dukkhanirodhagāminī patipadā ariyasaccaṃ bhāvetabbanti
me bhikkhave, pubbe ananussutesu dhammesu cakkhuṃ udapādi, ñāṇaṃ
udapādi, paññā udapādi, vijjā udapādi, āloko udapādi.

Taṃ kho panidaṃ dukkhanirodhagāminī patipadā ariyasaccaṃ bhāvitanti me
bhikkhave, pubbe ananussutesu dhammesu cakkhuṃ udapādi, ñāṇaṃ udapādi,
paññā udapādi, vijjā udapādi, āloko udapādi.

[Yāva kīvañca me] bhikkhave, imesu catūsu ariyasaccesu evantiparivaṭṭaṃ
dvādasākāraṃ yathābhūtaṃ ñāṇadassanaṃ na suvisuddhaṃ ahosi, neva tāvāhaṃ
bhikkhave, sadevake loke samārake sabrahmake sassamaṇabrāhmaṇiyā pajāya
sadevamanussāya anuttaraṃ sammāsambodhiṃ abhisambuddho paccaññāsiṃ.

Yato ca kho me bhikkhave, imesu catūsu ariyasaccesu evantiparivaṭṭaṃ
dvādasākāraṃ yathābhūtaṃ ñāṇadassanaṃ suvisuddham ahosi, athāham
bhikkhave, sadevake loke samārake sabrahmake sassamaṇabrāhmaṇiyā pajāya
sadevamanussāya anuttaraṃ sammāsambodhiṃ abhisambuddho paccaññāsiṃ.

Ñāṇañca pana me dassanaṃ udapādi, akuppā me vimutti ayamantimā jāti,
natthidāni punabbhavo ti.

% TODO: ? bhavagato

Idam avoca bhagavā. attamanā pañcavaggiyā bhikkhū bhāgavato bhāsitaṃ
abhinanduṃ.

\clearpage

\englishText
\markboth{\englishTitle}{\rightmark}

As this exposition was proceeding, the spotless, immaculate vision of
the Dhamma appeared to the Venerable Koṇḍañña and he knew: ‘Everything
that has the nature to arise has the nature to cease.’

When the Blessed One had set in motion the Wheel of Dhamma, the
Earthbound devas proclaimed with one voice,

‘The incomparable Wheel of Dhamma has been set in motion by the Blessed
One in the deer sanctuary at Isipatana, near Benares, and no seeker,
brahmin, celestial being, demon, god, or any other being in the world
can stop it.’

Having heard what the Earthbound devas said, the devas of the Four Great
Kings proclaimed with one voice\ldots

Having heard what the devas of the Four Great Kings said, the devas of
the Thirty-three proclaimed with one voice\ldots

Having heard what the devas of the Thirty-three said, the Yāma devas
proclaimed with one voice\ldots

Having heard what the Yāma devas said, the Devas of Delight proclaimed
with one voice\ldots

Having heard what the Devas of Delight said, the Devas Who Delight in
Creating, proclaimed with one voice\ldots

Having heard what the Devas Who Delight in Creating said, the Devas Who
Delight in the Creations of Others proclaimed with one voice\ldots

Having heard what the Devas Who Delight in the Creations of Others said,
the Brahma gods proclaimed in one voice,

‘The incomparable Wheel of Dhamma has been set in motion by the Blessed
One in the deer sanctuary at Isipatana, near Benares, and no seeker,
brahmin, celestial being, demon, god, or any other being in the world
can stop it.’

\clearpage

\paliText
\markboth{\paliTitle}{\rightmark}

Imasmiñca pana veyyākaraṇasmiṃ bhaññamāne āyasmato koṇḍaññassa virajaṃ
vītamalaṃ dhammacakkhuṃ udapādi: yaṅkiñci samudayadhammaṃ sabbantaṃ
nirodhadhamman ti.

[Pavattite ca bhagavatā] dhammacakke bhummā devā saddamanussāvesuṃ:

Etaṃ bhagavatā bārāṇasiyaṃ isipatane migadāye anuttaraṃ dhammacakkaṃ
pavattitaṃ appaṭivattiyaṃ samaṇena vā brāhmaṇena vā devena vā mārena vā
brahmunā vā kenaci vā lokasmin ti.

Bhummānaṃ devānaṃ saddaṃ sutvā, cātummahārājikā devā
saddamanussāvesuṃ\ldots

Cātummahārājikānaṃ devānaṃ saddaṃ sutvā, tāvatiṃsā devā
saddamanussāvesuṃ\ldots

Tāvatiṃsānaṃ devānaṃ saddaṃ sutvā, yāmā devā saddamanussāvesuṃ\ldots

Yāmānaṃ devānaṃ saddaṃ sutvā, tusitā devā saddamanussāvesuṃ\ldots

Tusitānaṃ devānaṃ saddaṃ sutvā, nimmānaratī devā saddamanussāvesum\ldots

Nimmānaratīnaṃ devānaṃ saddaṃ sutvā, paranimmitavasavattī devā
saddamanussāvesuṃ\ldots

Paranimmitavasavattīnaṃ devānaṃ saddaṃ sutvā, brahmakāyikā devā
saddamanussāvesuṃ:

Etaṃ bhagavatā bārāṇasiyaṃ isipatane migadāye anuttaraṃ dhammacakkaṃ
pavattitaṃ appaṭivattiyaṃ samaṇena vā brāhmaṇena vā devena vā mārena vā
brahmunā vā kenaci vā lokasmin ti.

\clearpage

\englishText
\markboth{\englishTitle}{\rightmark}

Thus in a moment, an instant, a flash, word of the Setting in Motion of
the Wheel of Dhamma went forth up to the Brahma world, and the
ten-thousandfold universal system trembled and quaked and shook, and a
boundless, sublime radiance surpassing the power of devas appeared on
earth.

Then the Blessed One made the utterance,

‘Truly, Koṇḍañña has understood, Koṇḍañña has understood!’ Thus it was
that the Venerable Koṇḍañña got the name Aññā-Koṇḍañña: ‘Koṇḍañña Who
Understands.’

Thus ends the discourse on Setting in Motion the Wheel of Dhamma.

\clearpage

\paliText
\markboth{\paliTitle}{\rightmark}

Itiha tena khaṇena, tena muhuttena, yāva brahmalokā saddo abbhuggacchi.
ayañca dasasahassī lokadhātu saṅkampi sampakampi sampavedhi, appamāṇo ca
oḷāro obhāso loke pāturahosi atikkammeva devānaṃ devānubhāvaṃ.

Atha kho bhagavā udānaṃ udānesi:

Aññāsi vata bho koṇḍañño, aññāsi vata bho koṇḍañño ti. itihidaṃ āyasmato
koṇḍaññassa aññākoṇḍañño tveva nāmaṃ ahosī ti.

Dhammacakkappavattana-suttaṃ niṭṭhitaṃ.

\chapterTocDelegatePageNumber
\chapter{The Characteristic of Not-Self}% {{{1

\setTocDelegatedPageNumber
\englishText
\renewcommand{\englishTitle}{The Characteristic of Not-Self}

\begin{leader}
\soloinstr{Solo introduction}

All beings should take pains to understand the characteristic of
not-self, which provides matchless deliverance from self-view and
self-perception, as taught by the supreme Buddha.

This teaching is given so that those who meditate on experienceable
realities may arrive at perfect comprehension;

It is for the development of perfect understanding of these phenomena,
and for the investigation of all defiled mind-moments.

The consequence of this practice is total deliverance, so, desirous of
bringing this teaching forth with its great benefit, let us now recite
this Sutta.

\end{leader}

[Thus have I heard]

At one time the Blessed One was dwelling at Benares in the deer park.
There he addressed the group of five bhikkhus:

‘Form, bhikkhus, is not-self. If, bhikkhus, form were self, then form
would not lead to affliction, and one might be able to say in regard to
form, “Let my form be thus, let my form not be thus.” But since,
bhikkhus, form is not-self, form therefore leads to affliction, and one
is not able to say in regard to form, “Let my form be thus, let my form
not be thus.”

‘Feeling is not-self. If, bhikkhus, feeling were self, feeling would
not lead to affliction, and one might be able to say in regard to
feeling, “Let my feeling be thus, let my feeling not be thus.” But
since, bhikkhus, feeling is not-self, feeling therefore leads to
affliction, and one is not able to say in regard to feeling, “Let my
feeling be thus, let my feeling not be thus.”

\chapterTocSubIndentTrue
\chapter{Anatta-lakkhaṇa Sutta}% {{{1

\paliText
\renewcommand{\paliTitle}{Anatta-lakkhaṇa Sutta}

\begin{leader}
\soloinstr{Solo introduction}

{\setlength{\tabcolsep}{0.9em}
\begin{solotwochants}
Yantaṃ sattehi dukkhena & ñeyyaṃ anattalakkhaṇaṃ\\
Attavādattasaññāṇaṃ  & sammadeva vimocanaṃ\\
Sambuddho taṃ pakāsesi & diṭṭhasaccāna yoginaṃ\\
Uttariṃ paṭivedhāya & bhāvetuṃ ñāṇamuttamaṃ\\
Yantesaṃ diṭṭhadhammānam & ñāṇenupaparikkhataṃ\\
Sabbāsavehi cittāni & vimucciṃsu asesato\\
Tathā ñāṇānussārena & sāsanaṃ kātumicchataṃ\\
Sādhūnaṃ atthasiddhatthaṃ & taṃ suttantaṃ bhaṇāma se.\\
\end{solotwochants}
}
\end{leader}

[Evaṃ me sutaṃ]

Ekaṃ samayaṃ bhagavā bārāṇasiyaṃ viharati isipatane migadāye. tatra kho
bhagavā pañcavaggiye bhikkhū āmantesi:

Rūpaṃ bhikkhave anattā, rūpañca hidaṃ bhikkhave attā abhavissa, nayidaṃ
rūpaṃ ābādhāya saṃvatteyya, labbhetha ca rūpe, evaṃ me rūpaṃ hotu, evaṃ
me rūpaṃ mā ahosī ti. yasmā ca kho bhikkhave rūpaṃ anattā, tasmā rūpaṃ
ābādhāya saṃvattati, na ca labbhati rūpe, evaṃ me rūpaṃ hotu, evaṃ me
rūpaṃ mā ahosī ti.

Vedanā anattā, vedanā ca hidaṃ bhikkhave attā abhavissa, nayidaṃ vedanā
ābādhāya saṃvatteyya, labbhetha ca vedanāya, evaṃ me vedanā hotu, evaṃ
me vedanā mā ahosī ti. yasmā ca kho bhikkhave vedanā anattā, tasmā
vedanā ābādhāya saṃvattati, na ca labbhati vedanāya, evaṃ me vedanā
hotu, evaṃ me vedanā mā ahosī ti.

\clearpage

\englishText
\markboth{\englishTitle}{\rightmark}

‘Perception is not-self. If, bhikkhus, perception were self, perception
would not lead to affliction, and one might be able to say in regard to
perception, “Let my perception be thus, let my perception not be thus.”
But since, bhikkhus, perception is not-self, perception therefore leads
to affliction, and one is not able to say in regard to perception, “Let
my perception be thus, let my perception not be thus.”

‘Mental formations are not-self. If, bhikkhus, mental formations were
self, mental formations would not lead to affliction, and one might be
able to say in regard to mental formations, “Let my mental formations be
thus, let my mental formations not be thus.” But since, bhikkhus, mental
formations are not-self, mental formations therefore lead to affliction,
and one is not able to say in regard to mental formations, “Let my
mental formations be thus, let my mental formations not be thus.”

‘Consciousness is not-self. If, bhikkhus, consciousness were self,
consciousness would not lead to affliction, and one might be able to say
in regard to consciousness, “Let my consciousness be thus, let my
consciousness not be thus.” But since, bhikkhus, consciousness is
not-self, consciousness therefore leads to affliction, and one is not
able to say in regard to consciousness, “Let my consciousness be thus,
let my consciousness not be thus.”

‘What do you think about this, bhikkhus? Is form permanent or
impermanent?’

‘Impermanent, Venerable Sir.’

‘But is that which is impermanent painful or pleasurable?’

‘Painful, Venerable Sir.’

‘But is it fit to consider that which is impermanent, painful, of a
nature to change, as “This is mine, I am this, this is my self”?’

‘It is not, Venerable Sir.’

\clearpage

\paliText
\markboth{\paliTitle}{\rightmark}

Saññā anattā, saññā ca hidaṃ bhikkhave attā abhavissa, nayidaṃ saññā
ābādhāya saṃvatteyya, labbhetha ca saññāya, evaṃ me saññā hotu, evaṃ me
saññā mā ahosī ti.

Yasmā ca kho bhikkhave saññā anattā, tasmā, saññā ābādhāya saṃvattati,
na ca labbhati saññāya, evaṃ me saññā hotu, evaṃ me saññā mā ahosī ti.

Saṅkhārā anattā, saṅkhārā ca hidaṃ bhikkhave attā abhavissaṃsu, nayidaṃ
saṅkhārā ābādhāya saṃvatteyyuṃ, labbhetha ca saṅkhāresu, evaṃ me
saṅkhārā hontu, evaṃ me saṅkhārā mā ahesun ti.

Yasmā ca kho bhikkhave saṅkhārā anattā, tasmā saṅkhārā ābādhāya
saṃvattanti, na ca labbhati saṅkhāresu, evaṃ me saṅkhārā hontu, evaṃ me
saṅkhārā mā ahesun ti.

Viññāṇaṃ anattā, viññāṇañca hidaṃ bhikkhave attā abhavissa, nayidaṃ
viññānam ābādhāya saṃvatteyya, labbhetha ca viññāne evaṃ me viññāṇaṃ
hotu, evaṃ me viññāṇaṃ mā ahosī ti.

Yasmā ca kho bhikkhave viññāṇaṃ anattā, tasmā viññāṇaṃ ābādhāya
saṃvattati, na ca labbhati viññāne, evaṃ me viññāṇaṃ hotu, evaṃ me
viññāṇaṃ mā ahosī ti.

% TODO: 'ti? See further on too.
[Taṃ kiṃ maññatha bhikkhave], rūpam niccaṃ vā aniccaṃ vāti?

Aniccaṃ bhante.

Yam panāniccaṃ, dukkhaṃ vā taṃ sukhaṃ vāti?

Dukkhaṃ bhante.

Yam panāniccaṃ dukkhaṃ viparināma-dhammaṃ, kallaṃ nu taṃ samanupassituṃ,
etaṃ mama, esohamasmi, eso me attā ti?

No hetaṃ bhante.

\clearpage

\englishText
\markboth{\englishTitle}{\rightmark}

‘What do you think about this, bhikkhus? Is feeling permanent or
impermanent?’

‘Impermanent, Venerable Sir.’

‘But is that which is impermanent painful or pleasurable?’

‘Painful, Venerable Sir.’

‘But is it fit to consider that which is impermanent, painful, of a
nature to change, as “This is mine, I am this, this is my self”?’

‘It is not, Venerable Sir.’

‘What do you think about this, bhikkhus? Is perception permanent or
impermanent?’

‘Impermanent, Venerable Sir.’

‘But is that which is impermanent painful or pleasurable?’

‘Painful, Venerable Sir.’

‘But is it fit to consider that which is impermanent, painful, of a
nature to change, as “This is mine, I am this, this is my self”?’

‘It is not, Venerable Sir.’

‘What do you think about this, bhikkhus? Are mental formations
permanent or impermanent?’

‘Impermanent, Venerable Sir.’

‘But is that which is impermanent painful or pleasurable?’

‘Painful, Venerable Sir.’

‘But is it fit to consider that which is impermanent, painful, of a
nature to change, as “This is mine, I am this, this is my self”?’

‘It is not, Venerable Sir.’

\clearpage

\paliText
\markboth{\paliTitle}{\rightmark}

Taṃ kiṃ maññatha bhikkhave, vedanā niccā vā aniccā vāti?

Aniccā bhante.

Yam panāniccaṃ, dukkhaṃ vā taṃ sukhaṃ vāti?

Dukkhaṃ bhante.

Yam panāniccaṃ dukkhaṃ viparināma-dhammaṃ, kallaṃ nu taṃ samanupassituṃ,
etaṃ mama, esohamasmi, eso me attā ti?

No hetaṃ bhante.

Taṃ kiṃ maññatha bhikkhave, saññā niccā vā aniccā vāti?

Aniccā bhante.

Yam panāniccaṃ, dukkhaṃ vā taṃ sukhaṃ vāti?

Dukkhaṃ bhante.

Yam panāniccaṃ dukkhaṃ viparināma-dhammaṃ, kallaṃ nu taṃ samanupassituṃ,
etaṃ mama, esohamasmi, eso me attā ti?

No hetaṃ bhante.

Taṃ kiṃ maññatha bhikkhave, saṅkhārā niccā vā aniccā vāti?

Aniccā bhante.

Yam panāniccaṃ, dukkhaṃ vā taṃ sukhaṃ vāti?

Dukkhaṃ bhante.

Yam panāniccaṃ dukkhaṃ viparināma-dhammaṃ, kallaṃ nu taṃ samanupassituṃ,
etaṃ mama, esohamasmi, eso me attā ti?

No hetaṃ bhante.

\clearpage

\englishText
\markboth{\englishTitle}{\rightmark}

‘What do you think about this, bhikkhus? Is consciousness permanent or
impermanent?’

‘Impermanent, Venerable Sir.’

‘But is that which is impermanent painful or pleasurable?’

‘Painful, Venerable Sir.’

‘But is it fit to consider that which is impermanent, painful, of a
nature to change, as “This is mine, I am this, this is my self”?’

‘It is not, Venerable Sir.’

‘Wherefore, bhikkhus, whatever form there is, past, future, present,
internal or external, gross or subtle, inferior or superior, whether it
is far or near, all form should, by means of right wisdom, be seen as it
really is, thus: “This is not mine, I am not this, this is not my self.”

‘Whatever feeling there is, past, future, present, internal or
external, gross or subtle, inferior or superior, whether it is far or
near, all feeling should, by means of right wisdom, be seen as it really
is, thus: “This is not mine, I am not this, this is not my self.”

‘Whatever perception there is, past, future, present, internal or
external, gross or subtle, inferior or superior, whether it is far or
near, all feeling should, by means of right wisdom, be seen as it really
is, thus: “This is not mine, I am not this, this is not my self.”

‘Whatever mental formations there are, past, future, present, internal
or external, gross or subtle, inferior or superior, whether they are far
or near, all mental formations should, by means of right wisdom, be seen
as they really are, thus: “This is not mine, I am not this, this is not
my self.”

\clearpage

\paliText
\markboth{\paliTitle}{\rightmark}

Taṃ kiṃ maññatha bhikkhave, viññāṇaṃ niccaṃ vā aniccaṃ vāti?

Aniccaṃ bhante.

Yam panāniccaṃ, dukkhaṃ vā taṃ sukhaṃ vāti?

Dukkhaṃ bhante.

Yam panāniccaṃ dukkhaṃ viparināma-dhammaṃ, kallaṃ nu taṃ samanupassituṃ
etaṃ mama, esohamasmi, eso me attā ti?

No hetaṃ bhante.

[Tasmā tiha bhikkhave] yaṅkiñci rūpaṃ atītānāgata-paccuppannaṃ ajjhattaṃ
vā bahiddhā vā oḷārikaṃ vā sukhumaṃ vā hīnaṃ vā paṇītaṃ vā yandūre
santike vā, sabbaṃ rūpaṃ netaṃ mama, nesohamasmi, na me so attā ti
evametaṃ yathābhūtaṃ sammappaññāya daṭṭhabbaṃ.

Yā kāci vedanā atītānāgata-paccuppannā ajjhattā vā bahiddhā vā oḷārikā
vā sukhumā vā hīnā vā paṇītā vā yā dūre santike vā, sabbā vedanā netaṃ
mama, nesohamasmi, na me so attā ti evametaṃ yathābhūtaṃ sammappaññāya
daṭṭhabbaṃ.

Yā kāci saññā atītānāgata-paccuppannā ajjhattā vā bahiddhā vā oḷārikā vā
sukhumā vā hīnā vā paṇītā vā yā dūre santike vā, sabbā saññā netaṃ mama,
nesohamasmi, na me so attā ti evametaṃ yathābhūtaṃ sammappaññāya
daṭṭhabbaṃ.

Ye keci saṅkhārā atītānāgata-paccuppannā ajjhattā vā bahiddhā vā oḷārikā
vā sukhumā vā hīnā vā paṇītā vā ye dūre santike vā, sabbe saṅkhārā netaṃ
mama, nesohamasmi, na me so attā ti evametaṃ yathābhūtaṃ sammappaññāya
daṭṭhabbaṃ.

\clearpage

\englishText
\markboth{\englishTitle}{\rightmark}

‘Whatever consciousness there is, past, future, present, internal or
external, gross or subtle, inferior or superior, whether far or near,
all consciousness should, by means of right wisdom, be seen as it really
is, thus: “This is not mine, I am not this, this is not my self.”

‘Seeing in this way, bhikkhus, the wise noble disciple becomes
disenchanted with form, becomes disenchanted with feeling, becomes
disenchanted with perception, becomes disenchanted with mental
formations, becomes disenchanted with consciousness. Becoming
disenchanted, their passions fade away; with the fading of passion the
heart is liberated; with liberation there comes the knowledge: “It is
liberated,” and they know: “Destroyed is birth, the Holy Life has been
lived out, done is what had to be done, there is no more coming into any
state of being.”\thinspace ’

Thus spoke the Blessed One. Delighted, the group of five bhikkhus
rejoiced in what the Blessed One had said. Moreover, while this discourse was
being delivered, the minds of the five bhikkhus were freed from the
defilements, through clinging no more.

Thus ends the discourse on The Characteristic of Not-self.

\clearpage

\paliText
\markboth{\paliTitle}{\rightmark}

Yaṅkiñci viññāṇaṃ atītānāgata-paccuppannaṃ ajjhattaṃ vā bahiddhā vā
oḷārikaṃ vā sukhumaṃ vā hīnaṃ vā paṇītaṃ vā yandūre santike vā, sabbaṃ
viññāṇaṃ netaṃ mama, nesohamasmi, na me so attā ti evametaṃ yathābhūtaṃ
sammappaññāya daṭṭhabbaṃ.

Evaṃ passaṃ bhikkhave sutvā ariyasāvako rūpasmim pi nibbindati, vedanāya
pi nibbindati, saññāya pi nibbindati, saṅkhāresu pi nibbindati,
viññāṇasmim pi nibbindati, nibbindaṃ virajjati, virāgā vimuccati,
vimuttasmiṃ vimuttam iti ñāṇaṃ hoti, khīṇā jāti, vusitaṃ brahmacariyaṃ,
kataṃ karaṇīyaṃ, nāparaṃ itthattāyā ti pajānātī ti.

% TODO: other [ ] ?

[Idamavoca bhagavā.] attamanā pañcavaggiyā bhikkhū bhagavato bhāsitaṃ
abhinanduṃ. imasmiñca pana veyyākaraṇasmiṃ bhaññamāne pañcavaggiyānaṃ
bhikkhūnaṃ anupādāya āsavehi cittāni vimucciṃsūti.

Anattalakkhaṇa-suttaṃ niṭṭhitaṃ.

\chapterTocDelegatePageNumber
\chapter{The Fire Sermon}% {{{1

\setTocDelegatedPageNumber
\englishText
\renewcommand{\englishTitle}{The Fire Sermon}

\begin{leader}
\soloinstr{Solo introduction}

With his skill in training the trainable, the All-transcendent Buddha,
lucid speaker, teacher of the highest knowledge,

He who expounds to the people the Dhamma and Vinaya that is fitting and
worthy, teaching with this wonderful parable about fire, meditators of
the highest skill;

He has liberated those who listen with the liberation that is utterly
complete, through true investigation, with wisdom and attention.

Let us now recite this Sutta which describes the characteristics of
dukkha.

\end{leader}

[Thus have I heard]

At one time the Blessed One was staying near Gayā at Gayā Head together
with a thousand bhikkhus. There the Blessed One addressed the bhikkhus
thus:

‘Bhikkhus, everything is burning. And what, bhikkhus, is everything
that is burning?

‘The eye, bhikkhus, is burning, forms are burning, eye consciousness is
burning, eye contact is burning, the feeling that arises from eye
contact, whether it is pleasant, painful, or neutral, that too is
burning. With what is it burning? I declare that it is burning with the
fires of passion, hatred, and delusion; it is burning with birth,
ageing, and death, with sorrow, lamentation, pain, grief, and despair.

\enlargethispage{2\baselineskip}

‘The ear is burning, sounds are burning, ear consciousness is burning,
ear contact is burning, the feeling that arises from ear contact,
whether it is pleasant, painful, or neutral, that too is burning. With
what is it burning? I declare that it is burning with the fires of
passion, hatred, and delusion; it is burning with birth, ageing, and
death, with sorrow, lamentation, pain, grief, and despair.

\chapterTocSubIndentTrue
\chapter{Āditta-pariyāya Sutta}% {{{1

\paliText
\renewcommand{\paliTitle}{Āditta-pariyāya Sutta}

\begin{leader}
\soloinstr{Solo introduction}

\begin{solotwochants}
Veneyyadamanopāye  & sabbaso pāramiṃ gato\\
Amoghavacano buddho & abhiññāyānusāsako\\
Ciṇṇānurūpato cāpi & dhammena vinayaṃ pajaṃ\\
Ciṇṇāggipāricariyānaṃ & sambojjhārahayoginaṃ\\
Yamādittapariyāyaṃ & desayanto manoharaṃ\\
Te sotāro vimocesi & asekkhāya vimuttiyā\\
Tathevopaparikkhāya & viññūṇaṃ sotumicchataṃ\\
Dukkhatālakkhaṇopāyaṃ & taṃ suttantaṃ bhaṇāma se.\\
\end{solotwochants}
\end{leader}

[Evaṃ me sutaṃ]

Ekaṃ samayaṃ bhagavā gayāyaṃ viharati gayāsīse saddhiṃ bhikkhusahassena.
tatra kho bhagavā bhikkhū āmantesi:

Sabbaṃ bhikkhave ādittaṃ. kiñca bhikkhave sabbaṃ ādittaṃ?

Cakkhuṃ bhikkhave ādittaṃ, rūpā ādittā, cakkhuviññāṇaṃ ādittaṃ,
cakkhusamphasso āditto, yampidaṃ cakkhusamphassapaccayā uppajjati
vedayitaṃ sukhaṃ vā dukkhaṃ vā adukkhamasukhaṃ vā tam pi ādittaṃ. kena
ādittaṃ? ādittaṃ rāgagginā dosagginā mohagginā, ādittaṃ jātiyā
jarāmaraṇena sokehi paridevehi dukkhehi domanassehi upāyāsehi ādittanti
vadāmi.

Sotaṃ ādittaṃ, saddā ādittā, sotaviññāṇaṃ ādittaṃ, sotasamphasso āditto,
yampidaṃ sotasamphassapaccayā uppajjati vedayitaṃ sukhaṃ vā dukkhaṃ vā
adukkhamasukhaṃ vā tam pi ādittaṃ. kena ādittaṃ? ādittaṃ rāgagginā
dosagginā mohagginā, ādittaṃ jātiyā jarāmaraṇena sokehi paridevehi
dukkhehi domanassehi upāyāsehi ādittanti vadāmi.

\clearpage

\englishText
\markboth{\englishTitle}{\rightmark}

‘The nose is burning, odours are burning, nose consciousness is burning,
nose contact is burning, the feeling that arises from nose contact,
whether it is pleasant, painful, or neutral, that too is burning. With
what is it burning? I declare that it is burning with the fires of
passion, hatred, and delusion; it is burning with birth, ageing, and
death, with sorrow, lamentation, pain, grief, and despair.

‘The tongue is burning, tastes are burning, tongue consciousness is
burning, tongue contact is burning, the feeling that arises from tongue
contact, whether it is pleasant, painful, or neutral, that too is
burning. With what is it burning? I declare that it is burning with the
fires of passion, hatred, and delusion; it is burning with birth,
ageing, and death, with sorrow, lamentation, pain, grief, and despair.

‘The body is burning, tangible objects are burning, body consciousness
is burning, body contact is burning, the feeling that arises from body
contact, whether it is pleasant, painful, or neutral, that too is
burning. With what is it burning? I declare that it is burning with the
fires of passion, hatred, and delusion; it is burning with birth,
ageing, and death, with sorrow, lamentation, pain, grief, and despair.

‘The mind is burning, mental states are burning, mind consciousness is
burning, mind contact is burning, the feeling that arises through mind
contact, whether it is pleasant, painful, or neutral, that too is
burning. With what is it burning? I declare that it is burning with the
fires of passion, hatred, and delusion; it is burning with birth,
ageing, and death, with sorrow, lamentation, pain, grief, and despair.

‘Seeing thus, bhikkhus, the wise noble disciple becomes disenchanted
with the eye and disenchanted with forms, disenchanted with eye
consciousness, disenchanted with eye contact, and the feeling that
arises from eye contact --- whether it is pleasant, painful, or
neutral --- that too they become disenchanted with.

\clearpage

\paliText
\markboth{\paliTitle}{\rightmark}

Ghānaṃ ādittaṃ, gandhā ādittā, ghānaviññāṇaṃ ādittaṃ, ghānasamphasso
āditto, yampidaṃ ghānasamphassapaccayā uppajjati vedayitaṃ sukhaṃ vā
dukkhaṃ vā adukkhamasukhaṃ vā tam pi ādittaṃ. kena ādittaṃ? ādittaṃ
rāgagginā dosagginā mohagginā, ādittaṃ jātiyā jarāmaraṇena sokehi
paridevehi dukkhehi domanassehi upāyāsehi ādittanti vadāmi.

Jivhā ādittā, rasā ādittā, jivhāviññāṇam ādittaṃ, jivhāsamphasso āditto,
yampidaṃ jivhāsamphassapaccayā uppajjati vedayitaṃ sukhaṃ vā dukkhaṃ vā
adukkhamasukhaṃ vā tam pi ādittaṃ. kena ādittaṃ? ādittaṃ rāgagginā
dosagginā mohagginā, ādittaṃ jātiyā jarāmaraṇena sokehi paridevehi
dukkhehi domanassehi upāyāsehi ādittanti vadāmi.

Kāyo āditto, phoṭṭhabbā ādittā, kāyaviññāṇaṃ ādittaṃ, kāyasamphasso
āditto, yampidaṃ kāyasamphassapaccayā uppajjati vedayitaṃ sukhaṃ vā
dukkhaṃ vā adukkhamasukhaṃ vā tam pi ādittaṃ. kena ādittaṃ? ādittaṃ
rāgagginā dosagginā mohagginā, ādittaṃ jātiyā jarāmaraṇena sokehi
paridevehi dukkhehi domanassehi upāyāsehi ādittanti vadāmi.

Mano āditto, dhammā ādittā, manoviññāṇaṃ ādittaṃ, manosamphasso āditto,
yampidaṃ manosamphassapaccayā uppajjati vedayitaṃ sukhaṃ vā dukkhaṃ vā
adukkhamasukhaṃ vā tam pi ādittaṃ. kena ādittaṃ? ādittaṃ rāgagginā
dosagginā mohagginā, ādittaṃ jātiyā jarāmaraṇena sokehi paridevehi
dukkhehi domanassehi upāyāsehi ādittanti vadāmi.

\enlargethispage{2\baselineskip}

[Evaṃ passaṃ] bhikkhave sutvā ariyasāvako cakkhusmiṃ pi nibbindati,
rūpesu pi nibbindati, cakkhuviññāṇe pi nibbindati, cakkhusamphassepi
nibbindati, yampidaṃ cakkhusamphassapaccayā uppajjati vedayitaṃ sukhaṃ
vā dukkhaṃ vā adukkhamasukhaṃ vā tasmiṃ pi nibbindati.

\clearpage

\englishText
\markboth{\englishTitle}{\rightmark}

‘They become disenchanted with the ear, disenchanted with sounds,
disenchanted with ear consciousness, disenchanted with ear contact, and
the feeling that arises from ear contact --- whether it is pleasant,
painful, or neutral --- that too they become disenchanted with.

‘They become disenchanted with the nose, disenchanted with odors,
disenchanted with nose consciousness, disenchanted with nose contact,
and the feeling that arises from nose contact --- whether it is pleasant,
painful, or neutral --- that too they become disenchanted with.

‘They become disenchanted with the tongue, disenchanted with tastes,
disenchanted with tongue consciousness, disenchanted with tongue
contact, and the feeling that arises from tongue contact --- whether it is
pleasant, painful, or neutral --- that too they become disenchanted with.

‘They become disenchanted with the body, disenchanted with tangible
objects, disenchanted with body consciousness, disenchanted with body
contact, and the feeling that arises from body contact --- whether it is
pleasant, painful, or neutral --- that too they become disenchanted with.

‘They become disenchanted with the mind, disenchanted with mental
states, disenchanted with mind consciousness, disenchanted with mind
contact, and the feeling that arises from mind contact --- whether it is
pleasant, painful, or neutral --- that too they become disenchanted with.

‘Becoming disenchanted, their passions fade away; with the fading of
passion the heart is liberated; with liberation there comes the
knowledge: “It is liberated,” and they know: “Destroyed is birth, the
Holy Life has been lived out, done is what had to be done, there is no
more coming into any state of being.”\thinspace ’

\enlargethispage{\baselineskip}

Thus spoke the Blessed One; delighted, the bhikkhus rejoiced in what the
Blessed One had said. Moreover, while this discourse was being uttered, the
minds of those thousand bhikkhus were freed from the defilements,
without any further attachment.

Thus ends The Fire Sermon.

\clearpage

\paliText
\markboth{\paliTitle}{\rightmark}

Sotasmiṃ pi nibbindati, saddesu pi nibbindati, sotaviññāṇe pi
nibbindati, sotasamphassepi nibbindati, yampidaṃ sotasamphassapaccayā
uppajjati vedayitaṃ sukhaṃ vā dukkhaṃ vā adukkhamasukhaṃ vā tasmiṃ pi
nibbindati.

Ghānasmiṃ pi nibbindati, gandhesu pi nibbindati, ghānaviññāṇe pi
nibbindati, ghānasamphassepi nibbindati, yampidaṃ ghānasamphassapaccayā
uppajjati vedayitaṃ sukhaṃ vā dukkhaṃ vā adukkhamasukhaṃ vā tasmiṃ pi
nibbindati.

Jivhāya pi nibbindati, rasesu pi nibbindati, jivhāviññāṇe pi nibbindati,
jivhāsamphassepi nibbindati, yampidaṃ jivhāsamphassapaccayā uppajjati
vedayitaṃ sukhaṃ vā dukkhaṃ vā adukkhamasukhaṃ vā tasmiṃ pi nibbindati.

Kāyasmiṃ pi nibbindati, phoṭṭhabbesu pi nibbindati, kāyaviññāṇe pi
nibbindati, kāyasamphassepi nibbindati, yampidaṃ kāyasamphassapaccayā
uppajjati vedayitaṃ sukhaṃ vā dukkhaṃ vā adukkhamasukhaṃ vā tasmiṃ pi
nibbindati.

Manasmiṃ pi nibbindati, dhammesu pi nibbindati, manoviññāṇe pi
nibbindati, manosamphasse pi nibbindati, yampidaṃ manosamphassapaccayā
uppajjati vedayitaṃ sukhaṃ vā dukkhaṃ vā adukkhamasukhaṃ vā tasmiṃ pi
nibbindati.

Nibbindaṃ virajjati, virāgā vimuccati, vimuttasmiṃ vimuttam iti ñāṇaṃ
hoti, khīṇā jāti, vusitaṃ brahmacariyaṃ, kataṃ karaṇīyaṃ, nāparaṃ
itthattāyā ti pajānātī ti.

\enlargethispage{\baselineskip}

[Idamavoca bhagavā.] attamanā te bhikkhū bhagavato bhāsitaṃ abhinanduṃ.
imasmiñca pana veyyākaraṇasmiṃ bhaññamāne tassa bhikkhusahassassa
anupādāya āsavehi cittāni vimucciṃsū ti.

Ādittapariyāya-suttaṃ niṭṭhitaṃ.

\resumeNormalText

% End of suttas.tex
