\chapter{Glosario de términos en Pāli }

\enlargethispage{2\baselineskip}

\begin{description}

Anattā Literally, `not-self,' i.e. impersonal, without individual
  essence; neither a person nor belonging to a person. One of the three
  characteristics of conditioned phenomena.

\item[Anicca] Transient, impermanent, unstable, having the nature to
  arise and pass away. One of the three characteristics of conditioned
  phenomena.

\item[Añjali] A gesture of respect. The palms of both hands join
  together directly in front of the chest, with the fingers aligned and
  pointing upwards.

\item[Arahaṁ/Arahant] Literally, ‘worthy one’ --- a term applied to all
  enlightened beings. As an epithet of the Buddha alone, `Lord' is used.

\item[Ariyapuggalā] ‘Noble Beings’ or ‘Noble Disciples’ --- there are
  eight kinds: those who are working on or who have achieved the four
  different stages of realization.

\item[Bhagavā] Bountiful, with good fortune --- when used as an epithet of
  the Buddha, `the Fortunate One,' `the Blessed One.'

\item[Bhikkhu] A Buddhist monk who lives as an alms mendicant, abiding
  by 227 training precepts that define a life of renunciation and
  simplicity.

\item[Brahmā] Celestial being; a god in one of the higher spiritual
  realms.

\item[Buddha] The Understanding One, the Awakened One, who knows things
  as they are; a potential in every human being. The historical Buddha,
  Siddhattha Gotama, lived and taught in India in the 5th century B.C.E.

\item[Deva] A celestial being. Less refined than a brahmā; as a deva is
  still in a sensual realm, albeit a very refined one.

\item[Dhamma] (Sanskrit: Dharma) The Teaching of the Buddha as contained
  in the scriptures; not dogmatic in character, but more like a raft or
  vehicle to convey the disciple to deliverance. Also, the Truth towards
  which that Teaching points; that which is beyond words, concepts or
  intellectual understanding. When written as ‘\emph{dhamma}’, i.e.
  with lower case `d', this refers to an ‘item’ or ‘thing’.

\item[Dukkha] Literally, ‘hard to bear’ --- dis-ease, restlessness of
  mind, anguish, conflict, unsatisfactoriness, discontent, stress,
  suffering. One of the three characteristics of conditioned phenomena.

\item[Factors of Awakening (bojjhaṅga)] 1.~mindfulness, 2.~investigation of truth, 3.~effort, 4.~rapture, 5.~tranquility, 6.~concentration, 7.~equanimity.

\item[Foundations of Mindfulness (satipaṭṭhāna)] Mindfulness of 1.~\emph{kāya} (body), 2.~\emph{vedanā} (feelings), 3.~\emph{citta} (mind),
4.~\emph{dhamma} (mind-objects).

\item[Grounds of Birth (yoni)] The four modes of generation by which
  beings take birth: womb-born, egg-born, moisture-born and spontaneously born.

\item[Holy Life (brahmacariya)] Literally: the Brahma-conduct; usually
  referring to the monastic life. Using this term emphasizes the vow of
  celibacy.

\item[Jhāna] Mental absorption. A state of strong concentration focused
  on a single physical or sensation or mental notion.

\item[Kamma] (Sanskrit: karma) Action, deed; actions created by habitual
  impulse, intention, volition.

\item[Khandhā] The five aggregates, physical or mental ---
  that is: \emph{rūpa, vedanā, saññā, saṅkhārā, viññāṇa.} Attachment to
  any of these as, ‘This is mine’, ‘I am this’ or, ‘This is my self’ is
  \emph{upādāna} --- clinging or grasping.

\item[Māra] Personification of evil forces. During the Buddha’s struggle
  for enlightenment, Māra manifested frightening and enticing forms to
  try to turn him back from his goal.

\item[Nibbāna] (Sanskrit: Nirvāṇa) Literally, ‘coolness’ --- the state of
  liberation from all suffering and defilements, the goal of the
  Buddhist path.

\item[Paccekabuddha] Solitary Buddha --- someone enlightened by their own
  efforts without relying on a teacher but who, unlike the Buddha, has
  no following of disciples.

\item[Paritta] Verses chanted particularly for blessing and protection.

\item[Parinibbāna] The Buddha’s final passing away, i.e. final entering
  Nibbāna.

\item[Peaceful Sage (muni)] An epithet of the Buddha.

\item[Planes of Birth (bhūmi)] The three planes where rebirth takes
  place: \emph{kāmāvacara-bhūmi}: the sensuous plane;
  \emph{rūpāvacara-bhūmi}: form plane; \emph{arūpāvacara-bhūmi}: formless
  plane.

\item[Puñña] Merit, the accumulation of good fortune, blessings, or
  well-being resulting from the practice of Dhamma.

\item[Rūpa] Form or matter. The physical elements that make up the body,
  i.e. earth, water, fire and air (solidity, cohesion, temperature and
  vibration).

\item[Saṅgha] The community of those who practise the Buddha’s Way.

  More specifically, those who have formally committed themselves to the
lifestyle of mendicant monks and nuns. The `four pairs, the eight kinds of
noble beings' are those who are on the path to or who have realized the
fruition of the four stages of enlightenment: stream-entry, once-return,
non-return and arahantship.

\item[Saṅkhārā] Formations, constructions, all conditioned things, or volitional
  impulses, that is all mental states apart from feeling and perception
  that colour one’s thoughts and make them either good, bad or neutral.

\item[Saññā] Perception, the mental function of recognition.

\item[Tathāgata] ‘Thus gone’ or ‘Thus come’ --- one who has gone beyond
  suffering and mortality; one who experiences things as they are,
  without delusion. The epithet that the Buddha applied to himself.

\item[Threefold bliss] Mundane bliss, celestial bliss and Nibbānic
  bliss.

\item[Triple Gem] Buddha, Dhamma and Saṅgha.

\item[Vedanā] Feeling --- physical and mental feelings that may be either
  pleasant, unpleasant or neutral.

\item[Viññāṇa] Sense consciousness --- the process whereby there is
  seeing, hearing, smelling, tasting, touching and thinking.

\end{description}

