\chapter{Dedication of offerings}

[Yo so] bh\cD{a}gavā \cD{a}rahaṃ sammāsambuddho

\begin{english}
To the Bl\cD{e}ssed One, th\cD{e} Lord, who f\cU{u}lly \cD{a}ttained perfect enl\cD{i}ghtenment,
\end{english}

Svākkh\cU{ā}to yena bh\cD{a}gav\cU{a}tā dhammo

\begin{english}
To th\cD{e} Teaching which he exp\cD{o}und\cD{e}d so well,
\end{english}

Supaṭ\cD{i}panno yassa bh\cD{a}gav\cD{a}to sāvak\cD{a}saṅgho

\begin{english}
And to th\cD{e} Blessed One's disc\cU{i}ples who have pr\cD{a}ctised well,
\end{english}

Tam-m\cU{a}yaṃ bh\cD{a}gavantaṃ s\cD{a}dhammaṃ s\cD{a}saṅghaṃ

\begin{english}
To these - th\cD{e} Buddha, th\cD{e} Dhamma, \cD{a}nd the S\cU{a}ngha -
\end{english}

Imeh\cU{i} sakkāreh\cD{i} yathārahaṃ ārop\cD{i}tehi \cD{a}bh\cU{i}pūj\cD{a}yāma

\begin{english}
We render w\cD{i}th offerings our r\cU{i}ghtful h\cDL{o}mage.
\end{english}

Sādh\cU{u} no bhante bh\cD{a}gavā s\cD{u}cira-parinibb\cD{u}topi

\begin{english}
It is w\cU{e}ll for us that the Bl\cD{e}ssed One, having attained l\cD{i}ber\cU{a}tion,
\end{english}

Pacch\cU{i}mā-j\cD{a}n\cU{a}tān\cU{u}kamp\cD{a}-mānasā

\begin{english}
Still had c\cD{o}mpassion for later g\cU{e}ner\cDL{a}tions.
\end{english}

Ime sakkāre dugg\cD{a}t\cD{a}-paṇṇākār\cU{a}-bhūte p\cD{a}ṭiggaṇhātu

\begin{english}
May th\cD{e}se simple of\cU{f}erings be acc\cD{e}pted
\end{english}

Amh\cU{ā}kaṃ dīgh\cD{a}rattaṃ h\cD{i}tāya s\cD{u}khāya

\begin{english}
For \cD{o}ur long-lasting benefit and f\cD{o}r the h\cU{a}ppiness it gi\cDL{v}es us.
\end{english}

Arahaṃ sammāsambuddho bh\cD{a}gavā

\begin{english}
Th\cD{e} Lord, th\cD{e} Perfectly Enl\cU{i}ghtened and Bl\cD{e}ssed One -
\end{english}

Buddhaṃ bh\cD{a}gavantaṃ \cD{a}bh\cU{i}vādemi

\begin{english}
\cD{I} render homage t\cD{o} the B\cU{u}ddha, the Bl\cD{e}ssed One.
\end{english}

\instr{Bow}

[Svākkh\cU{ā}to] bh\cD{a}gav\cU{a}tā dhammo

\begin{english}
Th\cD{e} Teaching so c\cD{o}mpletely expl\cU{a}ined by him --
\end{english}

Dhammaṃ namassāmi

\begin{english}
I b\cDL{o}w t\cD{o} th\cD{e} Dh\cD{a}mma.
\end{english}

\instr{Bow}

[Supaṭ\cD{i}panno] bh\cD{a}gav\cD{a}to sāvak\cD{a}saṅgho

\begin{english}
Th\cD{e} Blessed One's disc\cU{i}ples who have pr\cD{a}ctised well --
\end{english}

S\cU{a}ṅghaṃ n\cD{a}māmi

\begin{english}
I b\cDL{o}w t\cD{o} th\cD{e} S\cD{a}ngha.
\end{english}

\instr{Bow}

\chapter[Namo Tassa]{Preliminary homage}

\begin{leader}
  [H\cU{a}nda mayaṃ buddhass\cD{a} bh\cD{a}gavato pubbabhāga-namak\cD{ā}raṃ karomase]
\end{leader}

\begin{english}
  [N\cU{o}w let us pay preliminary homage to the B\cD{u}ddha.]
\end{english}

Namo tassa bh\cD{a}gav\cD{a}to arah\cD{a}to s\cU{a}mmās\cU{a}mbuddh\cD{a}ssa

\instr{Three times}

\begin{english}
  H\cU{o}m\cU{a}ge to th\cD{e} Bl\cD{e}ssed, N\cU{o}ble, \cD{a}nd P\cD{e}rfectly Enl\cU{i}ghtened One.

  \instr{Three times}
\end{english}

\chapter{Recollection of the Buddha}

% TODO: why is this 'Recollection'? It is buddhābhitthutiṃ, 'homage' as
% the others.

\begin{leader}
  [H\cU{a}nda mayaṃ buddhābh\cD{i}tth\cD{u}tiṃ karomase]
\end{leader}

\begin{english}
  [N\cU{o}w let us chant in praise of the B\cD{u}ddha.]
\end{english}

Yo so tath\cU{ā}g\cD{a}to \cD{a}rahaṃ sammāsambuddho

\begin{english}
  The Tathāg\cD{a}ta is the P\cU{u}re One, th\cD{e} Perfectly Enl\cU{i}ghtened One.
\end{english}

Vijjāc\cD{a}raṇ\cU{a}-sampanno

\begin{english}
  He is \cD{i}mpeccable \cD{i}n conduct and \cD{u}nderst\cU{a}nding,
\end{english}

S\cD{u}g\cD{a}to

\begin{english}
  The \cD{A}cc\cU{o}mplished One,
\end{english}

Lok\cD{a}v\cU{i}dū

\begin{english}
  Th\cD{e} Knower \cU{o}f th\cD{e} Worlds.
\end{english}

An\cU{u}tt\cD{a}ro puris\cD{a}damma-sārathi

\begin{english}
  He trains perfectly th\cD{o}se who w\cU{i}sh t\cU{o} b\cD{e} trained.
\end{english}

Satthā deva-m\cD{a}nussānaṃ

\begin{english}
  He is Teacher of g\cU{o}ds and h\cD{u}mans.
\end{english}

Buddho bh\cD{a}gavā

\begin{english}
  He is awake and h\cD{o}ly.
\end{english}

Yo imaṃ lokaṃ s\cD{a}devakaṃ s\cD{a}mārakaṃ s\cD{a}brahm\cD{a}kaṃ

\begin{english}
  In this world with \cD{i}ts gods, demons, \cD{a}nd kind sp\cU{i}rits,
\end{english}

Sass\cU{a}maṇa-brāhmaṇiṃ p\cD{a}jaṃ s\cD{a}deva-m\cD{a}nuss\cU{a}ṃ s\cD{a}yaṃ \cD{a}bhiññā sacchik\cD{a}t\cU{v}ā p\cD{a}vedesi

\begin{english}
  Its s\cU{e}ekers and s\cD{a}ges, c\cD{e}lestial and h\cD{u}m\cU{a}n beings, he has by \\d\cD{e}ep insight rev\cU{e}aled th\cD{e} Truth.
\end{english}

Yo dhammaṃ des\cU{e}si \cD{ā}d\cU{i}-kalyāṇaṃ majjh\cU{e}-k\cD{a}lyāṇaṃ \\p\cD{a}riyosāṇa-\cD{k}alyāṇaṃ

\begin{english}
  He has pointed out the Dh\cD{a}mma: beautiful in the beg\cD{i}nning, \\beautiful in the m\cU{i}ddle, beautiful \cD{i}n th\cU{e} end.
\end{english}

Sātth\cU{a}ṃ s\cD{a}byañjaṇaṃ kevala-p\cD{a}ripuṇṇaṃ p\cD{a}risuddhaṃ \\brahma-c\cD{a}r\cU{i}yaṃ p\cD{a}kāsesi

\begin{english}
  He has explained the Sp\cU{i}r\cU{i}t\cD{u}al Life of c\cD{o}mplete p\cU{u}rity in its \\\cU{e}ssence and conv\cD{e}ntions.
\end{english}

Tam-ah\cU{a}ṃ bh\cD{a}gavantaṃ \cD{a}bh\cU{i}pūj\cD{a}yāmi tam-ah\cU{a}ṃ bh\cD{a}gavantaṃ \\s\cD{i}ras\cU{ā} n\cD{a}māmi

\begin{english}
  I chant \cD{m}y praise t\cD{o} the Bl\cU{e}ssed One, I bow \cD{m}y head t\cD{o} \\th\cU{e} Bl\cU{e}ssed One.
\end{english}

\instr{Bow}

\chapter{Homage to the Dhamma}

\begin{leader}
  [H\cU{a}nda mayaṃ dhammābh\cD{i}tth\cD{u}tiṃ karomase]
\end{leader}

\begin{english}
[N\cU{o}w let us chant in praise of the Dh\cD{a}mma.]
\end{english}

Yo so svākkh\cU{ā}to bh\cD{a}gav\cU{a}tā dhammo

\begin{english}
  Th\cD{e} Dhamma is w\cD{e}ll exp\cU{o}unded by the Bl\cD{e}ssed One,
\end{english}

S\cU{a}ndiṭṭh\cD{i}ko

\begin{english}
  Apparent here \cD{a}nd now,
\end{english}

\cD{A}kāl\cD{i}ko

\begin{english}
  Timeless,
\end{english}

Eh\cD{i}pass\cD{i}ko

\begin{english}
  \cD{E}ncouraging inv\cD{e}stig\cU{a}tion,
\end{english}

Opanay\cD{i}ko

\begin{english}
  Leading \cD{i}nwards,
\end{english}

P\cD{a}cc\cD{a}ttaṃ ved\cU{i}t\cD{a}bbo viññūhi

\begin{english}
  To be \cD{e}xperienced ind\cD{i}vidually b\cU{y} th\cD{e} wise.
\end{english}

Tam-ah\cU{a}ṃ dhammaṃ \cD{a}bh\cU{i}pūj\cD{a}yāmi tam-ah\cU{a}ṃ dhammaṃ \\s\cD{i}ras\cU{ā} n\cD{a}māmi

\begin{english}
  I chant \cD{m}y praise t\cD{o} this T\cU{e}aching, I bow \cD{m}y head t\cD{o} th\cU{i}s Truth.
\end{english}

\instr{Bow}

\chapter{Homage to the Sangha}

\begin{leader}
  [H\cU{a}nda mayaṃ saṅghābh\cD{i}tth\cD{u}tiṃ karomase]
\end{leader}

\begin{english}
  [N\cU{o}w let us chant in praise of the S\cD{a}ngha.]
\end{english}

Yo so supaṭ\cD{i}panno bh\cD{a}gav\cD{a}to sāvak\cD{a}saṅgho

\begin{english}
  They are th\cD{e} Blessed One's disc\cU{i}ples, who have pr\cD{a}ctised well,
\end{english}

Ujupaṭ\cD{i}panno bh\cD{a}gav\cD{a}to sāvak\cD{a}saṅgho

\begin{english}
  Who have practised dir\cD{e}ctly,
\end{english}

Ñāyapaṭ\cD{i}panno bh\cD{a}gav\cD{a}to sāvak\cD{a}saṅgho

\begin{english}
  Who have practised ins\cU{i}ghtfully,
\end{english}

S\cU{ā}mīc\cD{i}p\cD{a}ṭ\cD{i}panno bh\cD{a}gav\cD{a}to sāvak\cD{a}saṅgho

\begin{english}
  Those who pr\cU{a}ctise with int\cD{e}grity --
\end{english}

Yadidaṃ cattāri puris\cD{a}yugāni aṭṭh\cU{a} puris\cD{a}pugg\cD{a}lā

\begin{english}
  That is the f\cD{o}ur pairs, th\cD{e} eight kinds of n\cU{o}bl\cD{e} beings --
\end{english}

Esa bh\cD{a}gav\cD{a}to sāvak\cD{a}saṅgho

\begin{english}
  These are th\cD{e} Blessed One's disc\cU{i}ples.
\end{english}

Āh\cD{u}ṇeyyo

\begin{english}
  Such ones \cD{a}re worthy of gifts,
\end{english}

Pāh\cD{u}ṇeyyo

\begin{english}
  Worthy of h\cD{o}spit\cU{a}lity,
\end{english}

Dakkh\cD{i}ṇeyyo

\begin{english}
  Worthy of \cD{o}fferings,
\end{english}

Añj\cD{a}li-k\cD{a}r\cU{a}ṇīyo

\begin{english}
  Worthy \cU{o}f r\cD{e}spect;
\end{english}

An\cU{u}tt\cD{a}raṃ puññakkh\cD{e}ttaṃ lokassa

\begin{english}
  They give \cD{o}ccasion for \cD{i}ncomparable g\cU{o}odness to ar\cD{i}se \\\cD{i}n the world.
\end{english}

Tam-ah\cU{a}ṃ saṅghaṃ \cD{a}bh\cU{i}pūj\cD{a}yāmi tam-ah\cU{a}ṃ saṅghaṃ \\s\cD{i}ras\cU{ā} n\cD{a}māmi

\begin{english}
  I chant \cD{m}y praise t\cD{o} this S\cU{a}ngha, I bow \cD{m}y head t\cD{o} \\th\cU{i}s S\cU{a}ngha.
\end{english}

\instr{Bow}

\chapter{Salutation to the Triple Gem}

\begin{leader}
  [H\cU{a}nda mayaṃ ratanattaya-paṇāma-gāth\cU{ā}yo ceva s\cU{a}ṃvega-\\parikittana-pāṭhañc\cD{a} bhaṇāmase]
\end{leader}

\begin{english}
  [N\cU{o}w let us chant our salutation to the Tr\cD{i}ple Gem and a passage \\to arouse \cU{u}rgency.]
\end{english}

Buddho s\cD{u}suddho k\cD{a}ruṇāmah\cU{a}ṇṇavo

\begin{english}
  The B\cD{u}ddha, absolute\cD{l}y pure, with \cU{o}cean-like comp\cD{a}ssion,
\end{english}

Yoccant\cD{a}-suddhabb\cD{a}ra-ñāṇ\cD{a}-loc\cD{a}no

\begin{english}
  Possessing th\cD{e} clear sight of w\cD{i}sdom,
\end{english}

Lokass\cD{a} pāpūp\cD{a}k\cU{i}les\cD{a}-ghāt\cD{a}ko

\begin{english}
  Destroyer \cD{o}f worldly self-corr\cU{u}ption
\end{english}

Vandām\cU{i} buddhaṃ \cD{a}h\cU{a}m-ād\cD{a}ren\cD{a} taṃ

\begin{english}
  Devot\cU{e}dly \cD{i}ndeed, that Buddha \cU{I} r\cD{e}vere.
\end{english}

Dhammo p\cD{a}dīpo v\cD{i}ya tass\cD{a} satth\cD{u}no

\begin{english}
  The Teaching of th\cD{e} Lord, like \cD{a} lamp,
\end{english}

Yo magg\cU{a}pākām\cD{a}t\cD{a}-bhed\cD{a}-bhinn\cD{a}ko

\begin{english}
  Illuminating th\cD{e} Path and its Fruit: the D\cD{e}athless,
\end{english}

Lokuttaro yo c\cD{a} t\cD{a}datth\cD{a}-dīp\cD{a}no

\begin{english}
  That which is bey\cU{o}nd the cond\cD{i}tioned world --
\end{english}

Vandām\cU{i} dhammaṃ \cD{a}h\cU{a}m-ād\cD{a}ren\cD{a} taṃ

\begin{english}
  Devot\cU{e}dly \cD{i}ndeed, that Dhamma \cU{I} r\cD{e}vere.
\end{english}

S\cU{a}ṅgho s\cD{u}khettābhyati-kh\cD{e}tta-s\cU{a}ññito

\begin{english}
  Th\cD{e} Sangha, the m\cD{o}st fertile gr\cU{o}und for cultiv\cD{a}tion,
\end{english}

Yo diṭṭh\cU{a}santo s\cD{u}g\cD{a}tān\cD{u}bodh\cD{a}ko

\begin{english}
  Those who have real\cD{i}sed peace, awakened after th\cD{e} \\Acc\cU{o}mplished One,
\end{english}

Lolapp\cD{a}hīno \cD{a}r\cU{i}yo s\cD{u}medh\cD{a}so

\begin{english}
  N\cU{o}ble \cD{a}nd wise, all longing ab\cD{a}ndoned --
\end{english}

Vandām\cU{i} saṅghaṃ \cD{a}h\cU{a}m-ād\cD{a}ren\cD{a} taṃ

\begin{english}
  Devot\cU{e}dly \cD{i}ndeed, that Sangha \cU{I} r\cD{e}vere.
\end{english}

Iccevam-ekant\cD{a}bh\cU{i}pūj\cD{a}-neyy\cD{a}kaṃ vatthuttayaṃ \\vand\cD{a}y\cD{a}tābh\cD{i}saṅkh\cD{a}taṃ

\begin{english}
  This salutation should b\cD{e} made to th\cU{a}t which is w\cD{o}rthy.
\end{english}

Puññaṃ m\cD{a}yā yaṃ m\cD{a}m\cD{a} sabb\cD{u}padd\cD{a}vā mā h\cU{o}nt\cD{u} ve tass\cD{a} p\cD{a}bhāv\cD{a}siddh\cD{i}yā

\begin{english}
  Through th\cD{e} power of s\cD{u}ch good a\cU{c}tion, may \cD{a}ll obstacles d\cU{i}s\cD{a}ppear.
\end{english}

Idha tath\cU{ā}g\cD{a}to loke \cD{u}ppanno \cD{a}rahaṃ sammāsambuddho

\begin{english}
  One who knows things as th\cD{e}y are has come int\cU{o} th\cD{i}s world; \\and he is an \cD{A}rahant, \cD{a} perfectly Aw\cU{a}kened being.
\end{english}

Dhammo c\cD{a} des\cD{i}to niyyān\cD{i}ko \cD{u}p\cD{a}s\cD{a}miko p\cD{a}rinibbān\cD{i}ko s\cU{a}mbodh\cD{a}gāmī s\cD{u}g\cD{a}tapp\cD{a}ved\cD{i}to

\begin{english}
  Purifying th\cD{e} way leading \cD{o}ut \cD{o}f d\cD{e}l\cD{u}sion, calming and d\cD{i}recting to p\cU{e}rfect peace, and leading to enl\cD{i}ghtenment --- this Way he h\cU{a}s m\cD{a}de known.
\end{english}

M\cU{a}yantaṃ dhammaṃ s\cD{u}tvā evaṃ jānāma

\begin{english}
  Having heard th\cD{e} Teaching, we kn\cD{o}w this:
\end{english}

Jātip\cD{i} dukkhā

\begin{english}
  Birth is d\cD{u}kkha,
\end{english}

Jarāp\cD{i} dukkhā

\begin{english}
  Ageing is d\cD{u}kkha,
\end{english}

M\cD{a}raṇamp\cD{i} dukkhaṃ

\begin{english}
  And death is d\cD{u}kkha;
\end{english}

S\cU{o}ka-p\cD{a}rideva-dukkh\cD{a}-domanass\cD{u}pāyās\cU{ā}p\cD{i} dukkhā

\begin{english}
  S\cU{o}rrow, lament\cD{a}tion, pain, grief, and d\cD{e}spair are d\cD{u}kkha;
\end{english}

Appiyeh\cD{i} s\cU{a}mp\cD{a}yogo dukkho

\begin{english}
  Assoc\cD{i}ation with th\cD{e} d\cD{i}sliked is d\cD{u}kkha;
\end{english}

Piyeh\cD{i} v\cU{i}pp\cD{a}yogo dukkho

\begin{english}
  Sep\cU{a}ration from th\cD{e} liked is d\cD{u}kkha;
\end{english}

Yampicch\cU{a}ṃ n\cD{a} labhati tamp\cD{i} dukkhaṃ

\begin{english}
  Not attaining one's w\cU{i}shes is d\cD{u}kkha.
\end{english}

S\cU{a}ṅkhittena pañc\cD{u}pādānakkh\cU{a}ndhā dukkhā

\begin{english}
  In brief, th\cD{e} five focuses of ide\cU{n}tity are d\cD{u}kkha.
\end{english}

Seyy\cD{a}thīdaṃ

\begin{english}
  These are as f\cD{o}llows:
\end{english}

Rūpūpādān\cD{a}kkh\cU{a}ndho

\begin{english}
  attachment \cD{t}o form,
\end{english}

Vedanūpādān\cD{a}kkh\cU{a}ndho

\begin{english}
  attachment to f\cU{e}eling,
\end{english}

S\cU{a}ññūpādān\cD{a}kkh\cU{a}ndho

\begin{english}
  attachment to perc\cD{e}ption,
\end{english}

S\cU{a}ṅkh\cU{ā}rūpādān\cD{a}kkh\cU{a}ndho

\begin{english}
  attachment to mental format\cU{i}ons,
\end{english}

Viññāṇūpādān\cD{a}kkh\cU{a}ndho

\begin{english}
  attachment t\cD{o} sense-c\cU{o}nsciousness.
\end{english}

Yesaṃ p\cD{a}riññāya

\begin{english}
  For th\cD{e} c\cD{o}mplete underst\cU{a}nding of this,
\end{english}

Dh\cD{a}ramāno s\cU{o} bh\cD{a}gavā evaṃ b\cD{a}hulaṃ s\cU{ā}v\cD{a}ke v\cD{i}neti

\begin{english}
  Th\cD{e} Blessed One i\cD{n} his l\cU{i}fetime frequently \cD{i}nstructed his disc\cU{i}ples \\in just th\cD{i}s way.
\end{english}

Evaṃ bhāgā c\cD{a} panassa bh\cD{a}gav\cD{a}to s\cU{ā}v\cD{a}kesu \cD{a}nus\cU{ā}s\cD{a}nī b\cD{a}hulā p\cD{a}vatt\cD{a}ti

\begin{english}
  In addition, he f\cD{u}rth\cD{e}r \cD{i}nstr\cD{u}cted:
\end{english}

Rūpaṃ \cD{a}niccaṃ

\begin{english}
  Form is imp\cD{e}rmanent,
\end{english}

Vedanā \cD{a}niccā

\begin{english}
  Feeling is imp\cD{e}rmanent,
\end{english}

S\cU{a}ññā \cD{a}niccā

\begin{english}
  Perc\cU{e}ption is imp\cD{e}rmanent,
\end{english}

S\cU{a}ṅkh\cU{ā}rā \cD{a}niccā

\begin{english}
  Mental formations are imp\cD{e}rmanent,
\end{english}

Viññāṇaṃ \cD{a}niccaṃ

\begin{english}
  Sense-c\cU{o}nsciousness is imp\cD{e}rmanent;
\end{english}

Rūpaṃ \cD{a}nattā

\begin{english}
  Form is n\cD{o}t-self,
\end{english}

Vedanā \cD{a}nattā

\begin{english}
  Feeling is n\cD{o}t-self,
\end{english}

S\cU{a}ññā \cD{a}nattā

\begin{english}
  Perc\cU{e}ption is n\cD{o}t-self,
\end{english}

S\cU{a}ṅkh\cU{ā}rā \cD{a}nattā

\begin{english}
  Mental formations are n\cD{o}t-self,
\end{english}

Viññāṇaṃ \cD{a}nattā

\begin{english}
  Sense-c\cU{o}nsciousness is n\cD{o}t-self;
\end{english}

S\cD{a}bbe s\cU{a}ṅkh\cU{ā}rā \cD{a}niccā

\begin{english}
  All conditions are \cD{t}ransient,
\end{english}

S\cD{a}bbe dhammā \cD{a}nattā'ti

\begin{english}
  There is n\cD{o} self in the cr\cD{e}ated or the \cD{u}ncr\cU{e}ated.
\end{english}

Te m\cU{a}yaṃ otiṇṇāmha-jāt\cD{i}yā j\cD{a}rāmaraṇena

\begin{english}
  All of us are bound \cD{b}y birth, ageing, \cD{a}nd death,
\end{english}

S\cU{o}keh\cD{i} p\cD{a}rideveh\cD{i} dukkh\cU{e}h\cD{i} domanasseh\cD{i} \cD{u}pāyāsehi

\begin{english}
  By s\cU{o}rrow, lament\cD{a}tion, pain, grief, and d\cD{e}spair,
\end{english}

Dukkh\cU{o}tiṇṇā dukkh\cD{a}p\cD{a}retā

\begin{english}
  B\cU{o}und by dukkha and obstr\cD{u}ct\cD{e}d \cD{b}y d\cD{u}kkha.
\end{english}

Appevanām\cU{i}mass\cD{a} kevalass\cD{a} dukkhakkh\cU{a}ndhass\cD{a} ant\cD{a}kir\cU{i}yā \\paññāyethā'ti

\begin{english}
  Let us all \cD{a}spire to c\cD{o}mplete fr\cU{e}edom from s\cD{u}ffering.
\end{english}

\instr{The following is chanted only by the monks and nuns}

C\cU{i}r\cU{a}par\cD{i}nibbutamp\cU{i} taṃ bh\cD{a}gav\cU{a}ntaṃ uddissa \cD{a}rah\cU{a}ntaṃ sammāsambuddhaṃ

\begin{english}
  Remembering the Bl\cD{e}ssed One, th\cD{e} Noble Lord, \cD{a}nd \\Perfectly Enl\cU{i}ghtened One, who long ago attained P\cD{a}r\cD{i}n\cD{i}bb\cD{ā}na,
\end{english}

Saddhā \cD{a}gārasmā anagār\cU{i}yaṃ pabb\cD{a}j\cD{i}tā

\begin{english}
  We have gone forth w\cD{i}th faith from home to h\cU{o}melessness,
\end{english}

Tasm\cU{i}ṃ bh\cD{a}gavati brahma-c\cD{a}r\cU{i}yaṃ c\cD{a}rāma

\begin{english}
  And like the Bl\cD{e}ssed One, we practise the H\cU{o}l\cD{y} Life,
\end{english}

Bhikkh\cU{ū}naṃ/Sīladhar\cU{ī}naṃ s\cU{i}kkhās\cD{ā}jīv\cD{a}-samāpannā

\begin{english}
  Being fully \cD{e}quipped with th\cD{e} bhikkhus'/nuns' s\cU{y}stem of tr\cD{a}ining.
\end{english}

Taṃ no brahma-c\cD{a}r\cU{i}yaṃ imass\cD{a} kevalass\cD{a} dukkhakkh\cU{a}ndhass\cD{a} ant\cD{a}kir\cU{i}yāya s\cU{a}ṃv\cU{a}tt\cD{a}tu

\begin{english}
  May this H\cD{o}ly Life lead us to th\cD{e} end of this wh\cU{o}le mass \\of s\cD{u}ffering.
\end{english}

\instr{An alternative version of the preceding section, which can be chanted by laypeople as well}

C\cU{i}r\cU{a}par\cD{i}nibbutamp\cU{i} taṃ bh\cD{a}gav\cU{a}ntaṃ saraṇaṃ g\cD{a}tā

\begin{english}
  The Bl\cD{e}ssed One, who long ago attained Parinibb\cU{ā}na, is our r\cD{e}fuge.
\end{english}

Dh\cU{a}mmañca S\cU{a}ṅghañca

\begin{english}
  So too are the Dh\cU{a}mma and the S\cD{a}ngha.
\end{english}

Tassa bh\cD{a}gavato s\cU{ā}sanaṃ yath\cU{ā}sati yath\cU{ā}balaṃ manasik\cD{a}roma \cD{a}nupaṭip\cU{a}jjāma

\begin{english}
  Attentively we f\cU{o}llow the pathway of that Bl\cD{e}ssed One, with all of \\our m\cU{i}ndfulness \cD{a}nd strength.
\end{english}

S\cU{ā} s\cU{ā} no p\cD{a}ṭ\cU{i}patti

\begin{english}
  May then the cultiv\cU{a}tion of this pr\cD{a}ctice
\end{english}

Imass\cD{a} kevalass\cD{a} dukkhakkh\cU{a}ndhass\cD{a} ant\cD{a}kir\cU{i}yāya s\cU{a}ṃv\cU{a}tt\cD{a}tu

\begin{english}
  Lead us to th\cD{e} end of ev\cU{e}ry kind of s\cD{u}ffering.
\end{english}

\chapter[Closing homage]{Closing homage}

[Arahaṃ] sammāsambuddho bh\cD{a}gavā

\begin{english}
  Th\cD{e} Lord, th\cD{e} Perfectly Enl\cU{i}ghtened and Bl\cD{e}ssed One -
\end{english}

Buddhaṃ bh\cD{a}gavantaṃ \cD{a}bh\cU{i}vādemi

\begin{english}
  \cD{I} render homage t\cD{o} the B\cU{u}ddha, the Bl\cD{e}ssed One.
\end{english}

\instr{Bow}

[Svākkh\cU{ā}to] bh\cD{a}gav\cU{a}tā dhammo

\begin{english}
  Th\cD{e} Teaching, so c\cD{o}mpletely expl\cU{a}ined by him --
\end{english}

Dhammaṃ namassāmi

\begin{english}
  I b\cDL{o}w t\cD{o} th\cD{e} Dh\cD{a}mma.
\end{english}

\instr{Bow}

[Supaṭ\cD{i}panno] bh\cD{a}gav\cD{a}to sāvak\cD{a}saṅgho

\begin{english}
  Th\cD{e} Blessed One's disc\cU{i}ples, who have pr\cD{a}ctised well --
\end{english}

S\cU{a}ṅghaṃ n\cD{a}māmi

\begin{english}
  I b\cDL{o}w t\cD{o} th\cD{e} S\cD{a}ngha
\end{english}

\instr{Bow}